---
title: 'Analysis #1'
author: "Sophia Drewry"
date: "11/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview: 
The goal of this analysis script is to create and fit non-time series models from quarterly averaged data. Quarterly averaging was done to account for the variable lag period between ENSO related weather events and the increases or decreases in dengue incidence. This variable lag period can be anywhere between 3-6 weeks, with warmer temperatures correlated with shorter lag periods and vice versa. For more background reasoning on the quarterly averaging rationale, please refer to the "ENSO and Vector Population" subsection in the 'Manuscript.Rmd' file. This quarterly averaging will result in fewer data points and thus a weaker model, but was done for the sake of producing models within the scope that the MADA class has covered. To see attempts at modeling weekly data with autoregressive models, please look at Analysis #2 script.

*HYPOTHESIS: Can monthly ENSO related weather variables predict monthly dengue case load?*


```{r}
library(readr)
library(dplyr) #for data processing
library(here) #to set paths
library(ggthemes)
library(RColorBrewer)
library(ggplot2)
library(ggcorrplot)
library(lubridate)
library(forcats)
library(zoo)
library(tidymodels)
library(glmnet)
library(parallel)
library(doParallel)
library(rpart)
library(rpart.plot) # to visualize DT plot
library(poissonreg)
library(MASS) # used for negative binomial regression
library(rcompanion) # for pulling and comparing stats from different models
library(kknn) # used for KKNN

#Load data
data_spot1 <- here::here("data","processed_data","Finaldata.rds")
FINALdta <- read_rds(data_spot1)
```

# Pre-processing
```{r}
str(FINALdta)
summary(FINALdta)
FINALdta$WeekDate <- as.Date(as.character(FINALdta$WeekDate)) 
# getting rid of unwanted variables: dengue serotype and other date variables
model.data <- FINALdta[c("WeekDate", "Total", "Estimated_population", "MinAT","MaxAT", "Precip", "TAvg", "SOI", "NINO4", "NINO3.4", "ENSO")]

barplot(tapply(FINALdta$Total, format(FINALdta$Month), FUN=sum)) #checking out how we want to split up the data. Traditional Jan-March will work since it is splitting up dengue peak season

# Checking out any N/A, using my favorite is.na line of code to see where the NA's are in various weather data locations. 
model.data  %>%  
  summarise_all(list(~is.na(.)))%>%
  pivot_longer(everything(), names_to = "variables", values_to="missing") %>%
  count(variables, missing)
## Only 4 found. Going to delete them because there are all on line 677
model.data <- model.data[-c(677), ]
```


### Splitting into quarterly data
As mentioned earlier, we are going to be changing this weekly time series data into quarterly.
```{r}
model.data$quarter = as.yearqtr(model.data$WeekDate)
qdta <- model.data %>% group_by(quarter) %>% summarise(
    Total = sum(Total), 
    Pop = mean(Estimated_population),
    MinAT = mean(MinAT),
    MaxAT = mean(MaxAT),
    Precip = mean(Precip),
    TAvg = mean(TAvg),
    SOI = mean(SOI),
    NINO4= mean(NINO4),
    NINO3.4= mean(NINO3.4),
    ENSO= sum(ENSO),) # Since it will be counted via quarter, this will in itself be a new function of ENSO month events/ quarter. 
barplot(tapply(qdta$Total, format(qdta$quarter), FUN=sum))

# Creating a correlation matrix  
A1P1tbl  <- cor(qdta[, unlist(lapply(qdta, is.numeric))])
A1P1tbl 
saveRDS(A1P1tbl, file = here("results","analysisfigures", "A1P1tbl.rds"))

# graphing correlation matrix  
A1P1 <- ggcorrplot(A1P1tbl, hc.order = TRUE, outline.col = "white")
A1P1
ggsave(filename = here("results", "analysisfigures", "A1P1.png"), plot = A1P1 )
# as expected, there is a correlation between weather variables. Keeping SOI values in mind

```
Covariates to keep in mind: 
* In General, ENSO and SST (3.4 & 4). For regular regression an SST location (3.4 or 4) will be chosen to be the best predictor. SST is used in determining ENSO
* SOI vs. ENSO, and SST 3.4 & SST 4 (- 0.8898684, -0.8202304, -0.80166581 respectively). For non variable selection models I am going to omit SOI
* TAvg vs. TMax (0.88823464). Going to omit TAvg in general because may be to many Air Temp predictors for non variable selection models

Setting data into time series form
```{r}
q.ts.dta  <- ts(qdta, start = c(2000), frequency = 4) # set to 4 to start the quarter
A1P2 <- plot.ts(qdta[,c(2:11)], main = "Quarterly Variable Plots") +  theme_minimal()  #Can only plot 10 variables with plot.ts, so I am taking out quarter
png(file = "results/analysisfigures/A1P2.png", width=600, height=400)

```

Checking to see if data is normally distributed
```{r}
which(is.na(model.data))
A1P3 <- hist(qdta$Total,col='red', main = "Hist: Quarterly Case Number Distribution") 
plot(A1P3)
A1P4 <- qqnorm(qdta$Total,col='red', main = "QQ Plot: Quarterly Case Number Distribution")
plot(A1P4)
# No, Dengue data is not normally distributed. Over dispersion
png(filename = here("results", "analysisfigures", "A1P3.png"))
png(filename = here("results", "analysisfigures", "A1P4.png"))

logqdta <- log(qdta$Total)
hist(logqdta, col='red', main = "Hist: Quarterly Case Number Distribution") 
```
Because data is overdispersed, going to try out both a Negative binomial regression and a Poisson regression with logged and non logged outcome

### Data splitting
Here we are going to split the data for testing and training models. 
- Training data will be used to fit the model. 2000/2001- 2008/2009 season  
- Testing set will be used to evaluate the model. 2009/2010- 2012/2013 season
* note that the date 2010 cuts off the 2008/2009 to the 2009/2010 season

```{r}
train.dta<- qdta %>%filter(quarter < 2010)
test.dta<- qdta %>%filter(quarter >= 2010)
```

--------------------------------------------------------------------------------
# Modeling
### Weather Predictors vs. Dengue Case Count
I am keeping the dependent variable as case counts because Poisson and negative binomial distribution require count data. To offset change in population I am including it as a variable. 

Weather predictors are as follows:
  From station data
- Population /// cont.
- MIN & MAT Air Temp  /// cont.
- Precipitation /// cont.
- Average Temp /// cont.
  From NOAA
- ENSO // categorical
- SOI /// cont.
- SST (region 3.4) /// cont.
 *See Pre-Processing section for more information on covariation* 

Now we are going to make the following models
* Null Model for comparison
* Multiple Linear Regression Model 
* Poisson Regression
* Negative binomial regression
* Decision Trees
* LASSO
* K Nearest Forest Model


Since there are colinear variables, I can expect the models with variable selection will provide the best model. 
--------------------------------------------------------------------------------
## Null Model
This is to use as a comparison for our other future models
```{r}
# creating model type
lm.mod <- linear_reg() %>% 
            set_engine('lm') %>% 
            set_mode('regression')

# Create null formula
n.rec <- recipe(Total ~ 1., data = train.dta)  

# Creating null recipe & model with TRAIN data
# set workflow
N.train.wflow <-
  workflow() %>% 
  add_model(lm.mod) %>% 
  add_recipe(n.rec)
# fitting
N.train.fit <- 
  N.train.wflow %>% 
  fit(data = train.dta)
augment(N.train.fit, train.dta) %>% yardstick::rmse(truth = Total, .pred)

# RMSE = 100.3227		
```

--------------------------------------------------------------------------------
## Multiple Linear Regression with PCA
Will be applying principal component analysis to address the multi-colinearity that exists between the variables.
Source: 
https://juliasilge.com/blog/cocktail-recipes-umap/
https://rdrr.io/github/tidymodels/learntidymodels/f/inst/tutorials/pca_recipes/pca_recipes.Rmd

```{r}
# creating model type
lm.mod <- linear_reg() %>% 
            set_engine('lm') %>% 
            set_mode('regression')

# create recipe
mlr.rec <- recipe(Total ~ Pop + MinAT + MaxAT + Precip + TAvg + SOI + NINO4 + NINO3.4 + ENSO,
  data = train.dta) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_normalize(SOI, ENSO) %>%
  step_center(all_predictors()) %>%
  step_pca(all_predictors()) # principal component analysis
```

### PCA
```{r}
mlr.rec # no extractions yet
pca.prep <- prep(mlr.rec) #training
tidy(pca.prep)
pca.loading <- tidy(pca.prep, 4)

# correlation between the principal component and the variable
A1ML.P1<- pca.loading %>%
  filter(component %in% paste0("PC", 1:5)) %>%
  mutate(component = fct_inorder(component)) %>%
  ggplot(aes(value, terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, nrow = 1) +
  labs(y = NULL) +
  theme_minimal() +
  scale_color_ptol("")
A1ML.P1
ggsave(filename = here("results", "analysisfigures", "A1ML.P1.png"), plot = A1ML.P1)


pca.bake <- bake(pca.prep, train.dta)

# Interested in PCA4 and PCA5. 
A1ML.P2 <- pca.bake %>%
  ggplot(aes(PC4, PC5, label=Total)) +
  geom_point(aes(color = Total), alpha = 0.7, size = 5)+
  geom_text(check_overlap = TRUE, hjust = "inward") +
  labs(color = NULL) +
 theme_minimal() 

A1ML.P2
ggsave(filename = here("results", "analysisfigures", "A1ML.P2.png"), plot = A1ML.P2)


# no obvious clustering 
```
Looks like PCA1 has to do mainly with population, not sure why it popped up. PCA 2-5 look the most interesting. Based on ENSO value, it looks like PCA2 and PCA3 describe "normal" weather phases, or neither El Nino or La Nina classified as 0 ENSO. PCA4 appears Cooler phases or La Nina cycle, and PCA5 seems do describe El Nino, however Precipitation is not very representative of this. 

```{r}
# set workflow
mlr.train.wflow <-
  workflow() %>% 
  add_model(lm.mod) %>% 
  add_recipe(mlr.rec)

# fitting
mlr.train.fit <- 
  mlr.train.wflow %>% 
  fit(data = train.dta)
augment(mlr.train.fit, train.dta) %>% yardstick::rmse(truth = Total, .pred)
# RMSE = 93.03668	
```


--------------------------------------------------------------------------------
## Poisson Regression
Sources:
https://poissonreg.tidymodels.org/reference/poisson_reg.html
https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/classification.html#poisson-regression
```{r}
# specifying the model using the poisson_reg() function
pois.spec <- poisson_reg() %>% 
  set_mode("regression") %>% 
  set_engine("glm")
#create dummy variable
pois.rec <- recipe(Total ~ quarter + Pop + MinAT + MaxAT + Precip + TAvg + SOI + NINO4 + NINO3.4 + ENSO,
  data = train.dta) %>% 
  step_dummy(all_nominal_predictors())
#workflow
pois.wf <- workflow() %>% 
  add_recipe(pois.rec) %>% 
  add_model(pois.spec)
#fit the model and look at the predictions.
pois.fit <- pois.wf %>% fit(data = train.dta)
A1PR.P1 <- augment(pois.fit, new_data = train.dta, type.predict = "response") %>% 
  ggplot(aes(Total, .pred)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, size = 1, color = "grey40") +
  labs(title = "Monthly Dengue Case numbers using Poission Regression ~ Training Data",
       x = "Actual", y = "Predicted")
A1PR.P1
ggsave(filename = here("results", "analysisfigures", "A1PR.P1.png"), plot = A1PR.P1)

augment(pois.fit, train.dta) %>% yardstick::rmse(truth = Total, .pred)

# RMSE = 77.94448	
```


--------------------------------------------------------------------------------

## Negative Binomial Regression

Here we are going to use the MASS package because tidymodels does not have a Negative Binomial Regression option. Here we are expecting regression coefficient is not significant even though many of these variables should be highly correlated with the predictor. For this reason, I am going to run multiple models with various predictors.
Sources:
https://data.princeton.edu/wws509/r/overdispersion
https://rcompanion.org/handbook/G_14.html

```{r}
# worried a bit about multicollinearity, So I am going to run a few models with and without ENSO related measures. Because of this reason variable selection will be important
summary(nbr1 <- glm.nb(Total ~  Pop + Precip +  TAvg + SOI, data = train.dta, maxit = 50)) # TAvg does not perform well
summary(nbr2 <- glm.nb(Total ~  Pop + Precip +  MinAT + MaxAT + SOI, data = train.dta, maxit = 50))
summary(nbr3 <- glm.nb(Total ~  Pop + Precip +  MinAT + MaxAT + SOI + NINO3.4, data = train.dta, maxit = 50))
summary(nbr4 <- glm.nb(Total ~  Pop + Precip +  MinAT + MaxAT + SOI + NINO4, data = train.dta, maxit = 50))
summary(nbr5 <- glm.nb(Total ~  Pop + Precip +  MinAT + MaxAT + NINO3.4, data = train.dta, maxit = 50))
# standard error is a bit large in all of these, to be expected
accuracy(list(nbr1, nbr2, nbr3, nbr4, nbr5),
          plotit=TRUE, digits=3)
```

Looks like best model is model 2: Total ~  Pop + Precip +  MinAT + MaxAT + SOI, had an RMSE = 101	


--------------------------------------------------------------------------------

## Decision Tree Model
Going to be using the Tidymodels framwork again. 

```{r}
folds <- vfold_cv(train.dta, v = 5, repeats = 5, strata = "Total")
# create recipe
dt.rec <- recipe(Total ~ Pop + MinAT + MaxAT + Precip + TAvg + SOI + NINO4 + NINO3.4 + ENSO,
  data = train.dta) %>% 
  step_dummy(all_nominal_predictors())

## Tuning hyperparameters
tune_spec <- 
  decision_tree(cost_complexity = tune(), 
  tree_depth = tune()) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")
tune_spec # We will come back to these parameters
# setting workflow
dt.wflow <- workflow() %>%
  add_model(tune_spec) %>%
  add_recipe(dt.rec)
```

### Tuning with a grid

```{r}
ncores = 4
cl <- makePSOCKcluster(ncores)
registerDoParallel(cl)
# Create a grid
dt.grid <- grid_regular(cost_complexity(),
                          tree_depth(),
                          levels = 5)
# tuning
tree.res <- dt.wflow %>% 
  tune_grid(resamples = folds, grid = dt.grid)
tree.res %>% collect_metrics()
stopCluster(cl)
```


### Visualization

```{r}
tree.res %>%
  collect_metrics() %>%
  mutate(tree_depth = factor(tree_depth)) %>%
  ggplot(aes(cost_complexity, mean, color = tree_depth)) +
  geom_line(size = 1.5, alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ .metric, scales = "free", nrow = 2) +
  scale_x_log10(labels = scales::label_number()) +
  scale_color_viridis_d(option = "plasma", begin = .9, end = 0)
# Looks like we have 2 deeper "trees" that perform similar in cost complexity as well, but not the best
# Lets check out the top 5
tree.res %>% show_best("rmse")
# Now to pull out the best set of hyperparameter values for our decision tree model
best.tree <- tree.res %>% select_best("rmse")
# finalize workflow
final.wf <- dt.wflow %>% finalize_workflow(best.tree)
# final fit
final.dt.fit <- final.wf %>% fit(data = train.dta) 
final.dt.fit
final.dt.pred <- predict(final.dt.fit, train.dta)
tree.res %>% show_best("rmse", n = 1)
# RMSE = 98.81927
```


## Visualize again

```{r}
rpart.plot(extract_fit_parsnip(final.dt.fit)$fit)
```

--------------------------------------------------------------------------------

## LASSO 

sources:
https://www.tidymodels.org/start/case-study/
https://stackoverflow.com/questions/66639452/tuning-a-lasso-model-and-predicting-using-tidymodels

### Building model

```{r}
folds <- vfold_cv(train.dta, v = 5, repeats = 5, strata = "Total")
# create recipe
lasso.rec <- recipe(Total ~ Pop + MinAT + MaxAT + Precip + TAvg + SOI + NINO4 + NINO3.4 + ENSO,
  data = train.dta) %>% 
  step_dummy(all_nominal_predictors())
 
# create model
lasso.mod <- linear_reg() %>%
  set_mode("regression") %>%           
  set_engine("glmnet") %>%
  set_args(penalty = tune(), mixture = 1)
# set workflow
lasso.wflow <- workflow() %>%
    add_model(lasso.mod) %>%
    add_recipe(lasso.rec)
```

### Train and tune LASSO

```{r}
### Setting cores
cores <- parallel::detectCores()
cores
ncores = 4
cl <- makePSOCKcluster(ncores)
registerDoParallel(cl)

# creating grid and tuning
lr_reg_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30)) 

# tuning on training data
lasso.res <- lasso.wflow %>% 
  tune::tune_grid(resamples = folds,
            grid = lr_reg_grid,
            control = control_grid(verbose = TRUE, save_pred = TRUE))
# turn off parallel cluster
lasso.res %>% autoplot()
stopCluster(cl)
```

### Choosing the best performing model

```{r}
lasso.top.models <- lasso.res %>% select_best("rmse") 
# finalize workflow with the best model
best.lasso.wflow <- lasso.wflow %>% 
  finalize_workflow(lasso.top.models)
# fitting best performing model
best.lasso.fit <- best.lasso.wflow %>% 
  fit(data = train.dta)
lasso.pred <- predict(best.lasso.fit, train.dta)
lasso.res %>% show_best(n = 1)
# RMSE = 112.85	

tidy(extract_fit_parsnip(best.lasso.fit)) %>% 
  filter(estimate != 0)

```
looks like the final variables were: Pop	MinAT	MaxAT	Precip SOI NINO4 ENSO

### Plotting performance

```{r}
# Variables and tuning perameters
x <- best.lasso.fit$fit$fit$fit
plot(x, "lambda")
```


--------------------------------------------------------------------------------

## K-Nearest Neighbors
As a general rule, k should equal sqrt of training set. In this case that is around 6
Sources: 
https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/classification.html#k-nearest-neighbors
https://www.r-bloggers.com/2021/02/using-tidymodels-to-predict-health-insurance-cost/
```{r}
# create recipe
knn.rec <- recipe(Total ~ quarter + Pop + MinAT + MaxAT + Precip + TAvg + SOI + NINO4 + NINO3.4 + ENSO,
  data = train.dta) %>% 
  step_dummy(all_nominal_predictors())

# create model
knn.mod <- nearest_neighbor() %>%
  step_log(Sale_Price, base = 10) %>% 
  set_mode("regression") %>%
  set_engine("kknn")

# set workflows for k = 1, 3 and 5
knn.wflow <- workflow() %>% add_recipe(knn.rec)
knn.wflow1 <- knn.wflow %>%  add_model(knn.mod %>% set_args(neighbors = 3))
knn.wflow2 <- knn.wflow %>%  add_model(knn.mod %>% set_args(neighbors = 6))
knn.wflow3 <- knn.wflow %>%  add_model(knn.mod %>% set_args(neighbors = 9))

# fit the models
knn.fit1 <- fit(knn.wflow1, data = train.dta)
knn.fit2 <- fit(knn.wflow2, data = train.dta)
knn.fit3 <- fit(knn.wflow3, data = train.dta)

augment(knn.fit1, train.dta) %>% yardstick::rmse(truth = Total, .pred)
augment(knn.fit2, train.dta) %>% yardstick::rmse(truth = Total, .pred)
augment(knn.fit3, train.dta) %>% yardstick::rmse(truth = Total, .pred)

summary(knn.fit2)
# RMSE = 52.77155	w/o resampling	
```

### Visualization on folds
```{r}
cores <- parallel::detectCores()
ncores = 4
cl <- makePSOCKcluster(ncores)
registerDoParallel(cl)
knn.cv <- vfold_cv(train.dta, prop = 0.9)
knn.rsmpl <- fit_resamples(knn.wflow2,
                           knn.cv,
                           control = control_resamples(save_pred = TRUE))

knn.rsmpl %>% collect_metrics()

# RMSE = 77.7879416		

knn.rsmpl %>%
    unnest(.predictions) %>%
    ggplot(aes(Total, .pred, color = id)) + 
    geom_abline(lty = 2, color = "gray80", size = 1.5) + 
    geom_point(alpha = 0.5) + 
   scale_color_ptol("") +
   theme_minimal() 

stopCluster(cl)
```





--------------------------------------------------------------------------------

# Final model

Model | RMSE
------------- | -------------
Null  Model | 100.3227	
Multiple Linear Regression Model | 93.03668	
Poisson Regression | 77.94448	
Negative binomial regression | 101
Decision Trees | 98.81927
LASSO | 113.3661	
K Nearest Neighbors Model | 77.7879416	

I am going to choose the Poisson Regression because this model was a bit more tuned compared to the K Nearest Forest Mode

```{r}

final.pois.fit <- pois.wf %>% fit(data = test.dta)

A1PR.P2 <- predict(final.pois.fit, new_data = test.dta)
augment(final.pois.fit, new_data = test.dta, type.predict = "response") %>% 
  ggplot(aes(Total, .pred)) +
  geom_point(alpha = 0.1 , size = 2) +
  geom_abline(slope = 1, size = 1, color = "grey40") +
  labs(title = "Monthly Dengue Case numbers using Poission Regression",
       x = "Actual", y = "Predicted") +
   scale_color_ptol("") +
   theme_minimal() 
A1PR.P2

final.pois.fit %>% 
  extract_fit_parsnip() %>% 
  tidy()

augment(final.pois.fit, test.dta) %>% yardstick::rmse(truth = Total, .pred)

# RMSE = 23.20935	

```




