---
title: "Exploratory Data Analysis"
author: "Sophia Drewry"
date: "10/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script loads performs an exploratory data analysis on data sets in the processed_data folder.
All results can be found in the Exploratory Data Analysis folder in the results folder 

load needed packages. make sure they are installed.

```{r}
library(readr)
library(dplyr) #for data processing
library(here) #to set paths
library(ggthemes)
library(RColorBrewer)
library(ggplot2)
library(lubridate)
library(ggforce)
library(table1)
library(tidymodels)

#Load data
data_spot1 <- here::here("data","processed_data","p.dengue.rds") 
data_spot2 <- here::here("data","processed_data","p.stationwk.rds")
data_spot3 <- here::here("data","processed_data","p.population.rds")
data_spot4 <- here::here("data","processed_data","p.NOAAdta.rds")
data_spot5 <- here::here("data","processed_data","Finaldata.rds")

#load data 
dengue <- read_rds(data_spot1)
stationwk <- read_rds(data_spot2)
population <- read_rds(data_spot3)
NOAAdta <- read_rds(data_spot4)
FINALdta <- read_rds(data_spot5)
```
# NOAA data  --------------------------------------------------------------------------------
## SST Values for region 4 and 3.4
```{r}
NP1 = ggplot(NOAAdta, aes(x, y), col = group) + 
  geom_line(data = NOAAdta, aes(x = Date, y = NINO4, color="NINO4")) +
  geom_line(data = NOAAdta, aes(x = Date, y = NINO3.4, color="NINO 3.4")) +
  xlab("Dates") +
  ylab("SST") +
  ggtitle("SST in Region 4 & 3.4 from 20") +
  scale_color_ptol("") +
  theme_minimal()
print(NP1)
Figure1 <- here("results", "exploratoryfigures", "EDAfig1.png")
ggsave(filename = Figure1, plot = NP1)
```

## SOI Values
```{r}
NP2 = ggplot() + 
  geom_line(data = NOAAdta, aes(x = Date, y = SOI)) +
  xlab("Dates") +
  ylab("SOI") +
  ggtitle("SOI Values") +
  scale_color_ptol("") +
  theme_minimal()

print(NP2)
Figure2 <- here("results", "exploratoryfigures", "EDAfig2.png")
ggsave(filename = Figure2, plot = NP2)
```

## SOI + SST
Ok, we know the two interact, so lets view them on the same graph
```{r}
coeff <- 27
NP3 = ggplot(NOAAdta, aes(x = Date), col = group) + 
  geom_line(data = NOAAdta, aes(x = Date, y = NINO4, color="NINO 4")) +
  geom_line(data = NOAAdta, aes(x = Date, y = NINO3.4, color="NINO 3.4")) +
  geom_line(data = NOAAdta, aes(x = Date, y = SOI + coeff, color = "SOI")) +
  xlab("Dates") +
  ylab("SST") +
  scale_color_ptol("") +
  theme_minimal() +
  scale_y_continuous(sec.axis = sec_axis(~. - coeff), name="SOI") +
  ggtitle("SOI & SST in Region 4 & 3.4 Over Time")
print(NP3)
Figure3 <- here("results", "exploratoryfigures", "EDAfig3.png")
ggsave(filename = Figure3, plot = NP3)
```

## Calendar view of El Nino/ La Nino monthly occurrence by year
```{r}
NP4 <- ggplot(NOAAdta, aes(x=Month, y=Year)) +
  geom_tile(color = "white",
            lwd = 1.5,
            linetype = 1, aes(fill = ENSO)) +
  labs(title = "El Nino/La Nina Occurance", y = "Year") +
  coord_fixed() +
  theme_minimal() +
  scale_fill_gradient2(low = "#205EAC", mid = "#d3d3d3", high = "#AE2033")

print(NP4)
Figure4 <- here("results", "exploratoryfigures", "EDAfig4.png")
ggsave(filename = Figure4, plot = NP4)
```
    
# Station Data --------------------------------------------------------------------------------
```{r}
SP1 <- ggplot(stationwk, aes(x, y), col = group) + 
    geom_line(data = stationwk, aes(x = Week, y = MinAT, color = "Min Air Temp")) +
    geom_line(data = stationwk, aes(x = Week, y = MaxAT, color = "Max Air Temp")) +
    geom_line(data = stationwk, aes(x = Week, y = TAvg, color = "Average Temperature")) +
    geom_col(data = stationwk, aes(x = Week, y = Precip, color = "Precipitation")) +
    scale_y_continuous(sec.axis = sec_axis(~.), name="Celcuis") +
    xlab("Week") +
    ylab("Centimeters") +
    ggtitle("Weather Data by Week in Iquitos, Peru") +
    theme_minimal() +
    scale_color_ptol("")

print(SP1)
Figure5 <- here("results", "exploratoryfigures", "EDAfig5.png")
ggsave(filename = Figure5, plot = SP1)
```

            
# Dengue data  --------------------------------------------------------------------------------
```{r}
str(dengue)
dengue$WeekDate <- as.Date(dengue$WeekDate) 
# Weekly Reported Cases by Serotype
DP1 = ggplot(dengue, aes(x, y), col = group) + 
    geom_col(data = dengue, aes(x = WeekDate, y = Denv1, color = "Denv 1")) +
    geom_col(data = dengue, aes(x = WeekDate, y = Denv2, color = "Denv 2")) +
    geom_col(data = dengue, aes(x = WeekDate, y = Denv3, color = "Denv 3")) +
    geom_col(data = dengue, aes(x = WeekDate, y = Denv4, color = "Denv 4")) +
    geom_col(data = dengue, aes(x = WeekDate, y = Other, color = "Other")) +
    xlab("Week") +
    ylab("Number of reported cases") +
    ggtitle("Weekly Reported Cases by Serotype in Iquitos Peru") +
    theme_minimal() +
    scale_color_ptol("")
print(DP1)
Figure6 <- here("results", "exploratoryfigures", "EDAfig6.png")
ggsave(filename = Figure6, plot = DP1)
```
## Cumulative incidence by season
```{r}
DP2 <- dengue %>%
  ggplot(aes(x = SeasonWk, y = SeasonCumCases)) +
  geom_point(color = "darkorchid4") +
  facet_wrap( ~ Season, ncol = 3) +
  labs(title = "Cumulative Incidence for Dengue Cases by Season in Iquitos, Peru",
       subtitle = "Data plotted by year",
       y = "# of Reported Cases",
       x = "Week") +
  theme_minimal() +
  scale_color_ptol("")
print(DP2)
Figure7 <- here("results", "exploratoryfigures", "EDAfig7.png")
ggsave(filename = Figure7, plot = DP2)
```
## Dengue Serotype Table
I want to get some yearly tables, so lets spit up the date by table
```{r}
dengue<- dengue %>% dplyr::mutate(year = lubridate::year(WeekDate), 
                        month = lubridate::month(WeekDate), 
                        day = lubridate::day(WeekDate))

table1::label(dengue$Denv1) <- "Denv1"
table1::label(dengue$Denv2) <- "Denv2"
table1::label(dengue$Denv3) <- "Denv3"
table1::label(dengue$Denv4) <- "Denv4"
table1::label(dengue$Total) <- "Total Cases"
Table1 <- table1::table1(~Denv1 + Denv2 + Denv3 + Denv4 + Total | year, data = dengue)
Table1

#save data frame table to file for later use in manuscript
Table1 = here("results","exploratoryfigures", "Table1.Rds")
saveRDS(Table1, file = Table1)
```

##############################################################################################  


# EDA With FINALdta --------------------------------------------------------------------------------
Some more visualization across datasets
```{r}
# Dengue Incidence Rate vs. Average Temperature
CP1 <- FINALdta %>% 
  ggplot(aes(x = IR, y = TAvg)) + 
  geom_point() + labs(title = "Dengue Incidence Rate vs. Average Temperature", x = "Incidence Rate per 100,000", y = "Average Temperature") +
  theme_minimal() +
  scale_color_ptol("")
plot(CP1)
ggsave(filename = here("results", "exploratoryfigures", "EDAfig8.png"), plot = CP1)

# Dengue Incidence Rate vs. Precip
CP2 <- FINALdta %>% 
  ggplot(aes(x = IR, y = Precip)) + 
  geom_point() + labs(title = "Dengue Incidence Rate vs. Precip", x = "Incidence Rate per 100,000", y = "Precip") +
  theme_minimal() +
  scale_color_ptol("")
plot(CP2)
ggsave(filename = here("results", "exploratoryfigures", "EDAfig9.png"), plot = CP2)

# Weather data by Season
CP3 <- ggplot(FINALdta, aes(x, y), col = group) + 
    geom_line(data = FINALdta, aes(x = SeasonWk, y = MinAT, color = "Min Air Temp")) +
    geom_line(data = FINALdta, aes(x = SeasonWk, y = MaxAT, color = "Max Air Temp")) +
    geom_line(data = FINALdta, aes(x = SeasonWk, y = TAvg, color = "Average Temperature")) +
    geom_col(data = FINALdta, aes(x = SeasonWk, y = Precip, color = "Precipitation")) +
    scale_y_continuous(sec.axis = sec_axis(~.), name="Celcuis") +
    facet_wrap( ~ Season, ncol = 3) +
    xlab("Week") +
    ylab("Centimeters") +
    ggtitle("Weather Data by Week in Iquitos, Peru: Stratified by Month") +
    theme_minimal() +
    scale_color_ptol("")
print(CP3)
ggsave(filename = here("results", "exploratoryfigures", "EDAfig10.png"), plot = CP3)

```
We have a couple ambiguous predictors such as SST region and Min and Max AT
## Choosing what SST NINO region (3.4 or 4) is the best predictor for Dengue IR
Running quick LM to determine which is a better predictor of IR. Both continuous variables
We will put both in the model
```{r}
# model 1: Total dengue vs. SST from 3.4
D3.4 <- lm(IR ~ NINO3.4, data = FINALdta)
# model 2: Total dengue vs. SST from 4
D4 <- lm(IR ~  NINO4, data = FINALdta)
# Compare both
anova(D3.4, D4) #Not sure if I should look at them separately in each model or run them together

# Model 3: Total dengue vs. SST from 3.4 & 4 
Dboth <- glm(IR ~ NINO4 + NINO3.4, data = FINALdta)
EAM1 <- summary(Dboth)
EAM1= here("results","exploratoryfigures", "EAM1.Rds")
saveRDS(EAM1, file = EAM1)

```
It appears NINO 3.4 is the best predictor.

## Choosing is min or max AirTemp is the best predictor for Dengue IR
Running quick LM to determine which is a better predictor of IR. Both continuous variables
```{r}
# model 1: Total dengue vs. MIN AT
MINat <- lm(IR ~ MinAT, data = FINALdta)
summary(MINat)
# model 2: Total dengue vs. MAX AT
MAXat <- lm(IR ~ MaxAT, data = FINALdta)
summary(MAXat)
# Compare both
anova(MAXat, MINat) #Not sure if I should look at them separately in each model or run them together

# Model 3: Total dengue vs. both MIN and MAX AT
ATboth <- glm(IR ~ MaxAT + MinAT, data = FINALdta)
summary(ATboth)
```



Both variables look pretty good. Min seems to perform better but it may be worth including both in the model